---
title: "Estimating Time on Task"
author: "Aleksander Dietrichson, PhD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calculating-time-on-task}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The learning analytics practitioner will often find it useful to have at their disposal a measure of how much time a student or other participant in a pedagogical process has spent interacting with an activity, course or another specific area. While millisecond precision measurements are unlikely absent the context of a controlled experimental design, we can probably convince ourselves that we can estimate *time-on-task* or *dwelltime*, by analyzing log-data, even if these data were created for a different purpose.



## Definition

One way of estimatig dwelltime $\hat{d}$ is: 

$$
\hat{d}  =    \sum_{i=1}^{|X|} 
\begin{cases}
    x_{i-1}-x_i,& \text{if } x_{i-1}-x_i\leq \theta\\
    0,              & \text{otherwise}
\end{cases}
$$

Where:

* $X$ is a series of log-entries sorted in descending order by their timestamp,
* $i$ is the index,
* $x_i$ denotes the individual items in the series $X$, and
* $\theta$ is the timeout parameter (i.e. max idle time before we assume a new session).

## Why timeout?

When users interact with Moodle they need to log in or otherwise start a session. While they are actively browsing the site they will create log-entries, however, we cannot guarantee that they log out --in fact our experience tells us that they do not. We therefore need to determine how to deal with the *last click*, and the path of least resistance is to discard this particular observation. Other options include imputing the data-point from the researchers favorite measure of central tendency. 


## Implementation with moodleR

Let's start by loading the `moodleR` package:

```{r, message=FALSE}
library(moodleR)
```

We will also be depending on dplyr verbs so:
```{r, message=FALSE}
library(dplyr)
```



```{r}
my_log <- mdl_log()
colnames(my_log)
```

We see that we have a 'timecreated` column and `userid` which can use to estimate *dwelltime*.

### Implementation for one user

```{r, include=FALSE}
set.seed(1617201903) #For reproducibility
```


First let's get a random user:

```{r}
my_user_id <- mdl_log() %>% 
  head(100000) %>% 
  slice_sample(n = 1) %>% 
  pull(userid)
```

And implement with dplyr verbs:

```{r}
dwelltime_estimate <- mdl_log() %>% 
  filter(userid == my_user_id) %>% 
  mutate(dwelltime =  timecreated - lag(timecreated))

mean(dwelltime_estimate %>% pull("dwelltime"), na.rm = TRUE) 
 
```

### Estimating timeout visually

Thus far we have implemented the algorithm without a $\theta$. This will very likely yield distorded values because it includes differences that span several sessions and time intervals which can be years. Let's take a look at the distribution to ascertain a reasonable value for our $theta$. For this we will use the ggplot2 package.


```{r}
library(ggplot2)
density_plot <-
dwelltime_estimate %>% 
  ggplot(aes(dwelltime)) +
  geom_density()+
  scale_x_log10(labels = scales::label_number_si())
density_plot
```


We see a flattening occurring  around 1000. Perhaps we can round this to 1200 (which corresponds to 20 minutes) for good measure:
```{r}
density_plot + 
  geom_vline(xintercept = 1200, lty=2, color="blue")
```

So,

```{r}
my_theta <- 1200
```


```{r}
dwelltime_estimate %>% 
  filter(dwelltime <=my_theta) %>% 
  summarize(total_dwelltime = sum(dwelltime, na.rm = TRUE))
```
And human readable: 

```{r}
dwelltime_estimate %>% 
  filter(dwelltime <=my_theta) %>% 
  summarize(total_dwelltime = sum(dwelltime, na.rm = TRUE)) %>% 
  pull(total_dwelltime) %>% 
  prettyunits::pretty_sec(compact = TRUE)
```

## Creating a timeline

It may be of use for the learning analytics researcher to now how the dwelltime is distributed over time. This can be accomplished by converting the `timecreated` column to a date-object using the [anytime](http://dirk.eddelbuettel.com/code/anytime.html) package.

```{r}
dwelltime_estimate <- dwelltime_estimate %>%
  filter(dwelltime <= my_theta) %>% 
  collect() %>% # Since we're using non-dplyr verbs we need to collect
  mutate(date = anytime::anydate(.data$timecreated)) %>% 
  group_by(date) %>% 
  summarize(
    dwelltime = sum(dwelltime, na.rm = TRUE)
    )


dwelltime_estimate %>% 
  ggplot(aes(x = date, y = dwelltime)) +
  geom_line()+
  scale_y_continuous(labels = prettyunits::pretty_sec, breaks = 1800*1:10)

```

Alternatively we can summarize by week using e.g. [lubridate](https://lubridate.tidyverse.org/):

```{r}
dwelltime_estimate %>% 
  mutate(Week = lubridate::week(date)) %>% 
  group_by(Week) %>% 
  summarize(
    dwelltime = sum(dwelltime, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = Week, y = dwelltime))+
  geom_line() +
  scale_y_continuous(labels = prettyunits::pretty_sec, breaks = c(1800*1:10))+
  scale_x_continuous( breaks = seq(0,53, by=5) )
```



