---
title: "Getting Started with moodleR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the moodleR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This vignette assumes that you have already been able to set up the connection to a Moodle Database and created a local cache (if chosen).

```{r setup}
library(moodleR)
library(ggplot2)
```

## Data access functions

Learning analytics practitioners working with LMS data will often be interested in a sub-set of the available tables with relevant references added. These can be accessed through several mdl_* functions which will connect to the relevant table(s) and return a reference to the data. Currently the package supports curated access through the following functions:

* `mdl_courses()`
* `mdl_forum_posts()`
* `mdl_grades()`
* `mdl_log()`
* `mdl_users()`


For example:

```{r, message=FALSE}
my_grades <- mdl_grades()
```

creates an R object named `my_grades` containing the dbplyr reference to the table in question. 

## Summary functions

The R generic `summary` function is implemented with sensible defaults so that:

```{r}
summary(my_grades)
```

provides useful information for the learning analytics preactitioner.

## Plot functions

Generic plot functions are also provided for all data-access objects:

```{r}
plot(my_grades)
```

The plot functions return a ggplot by default. If you prefer base graphics you can call the function thus:

```{r}
plot(my_grades, use_base_graphics = TRUE)
```


## Additional columns

For convenience some additional columns have been added to the tables. These are all snake_case and are thus easy to spot:

```{r}
library(dplyr)
my_grades %>% 
  select(contains("_")) %>% 
  colnames()
```

## Objects of class dbplyr

The return value from the data-access functions are also dbplyr tables so dbplyr verbs and operation work on them as well. For example:

```{r}
my_grades %>% 
  filter(courseid == 4957) %>% 
  summary()
```

```{r, warning=FALSE}
my_grades %>% 
  filter(courseid == 4957) %>% 
  plot()
```

## Using ggplot objects

Since the return value of a plot function is a ggplot object it is of course possible to add and/or manipulate the plot using the syntax from the ggplot2 package. For example by annotating the plot:

```{r}
my_plot <- mdl_grades() %>% 
  plot()
my_plot +
  annotate("text",.45,10^5, label="Passing Grade",cex=5)+
  annotate("segment",x=.6,y=10^5,xend=.6,yend=0, lty=2)
  
  
```


