---
title: "Getting Started with moodleR"
author: "Aleksander Dietrichson, PhD ( $\\chi^\\color{red}2$ Laboratories)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the moodleR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The moodleR package contains functions and algorithms to access Moodle$^{TM}$ data and perform reporting and/or learning analytics based in these data. For more information about Moodle go to [www.moodle.org](http://www.moodle.org). For more information about $\chi^\color{red}2$ (chi square) Laboratories and the work we do in learning analytics, visit our website: [www.chi2labs.com](http://www.chi2labs.com).

## Installing and loadng the package

To install the latest stable release from CRAN use:

```{r setup, eval=FALSE}
install.packages("moodleR")
```

To install the latest development version:
```{r, eval=FALSE}
devtools::install("chi2labs/moodleR")
```


After which the package is attached in the usual way.
```{r}
library(moodleR)
```

## Connecting to a Moodle Database


```{r, eval=FALSE}
mdl_con <- mdl_get_connection()
```


## Caching Data

There are several reasons you would want to cache data locally rather than operating directly on a moodle database:

* You don't want to create additional server traffic that will worry the system administrator and generate bandwidth and hardware charges.
* You are experimenting with exotic new features and want to avoid any accidents which may get you banned from accessing the data in the future
* You only need a subset of courses
* You are conducting exploratory research a sample of courses and want to make sure you data stays stable during your experimentation
* You want your Notebooks and Rmarkdown document to compile faster

The moodleR package provides a method `moodle_cache` which allows for local caching of the data. 
